Abundance/diversity GAM

### 2021-03-17 code/ms1_abundance.Rmd
update

# 1. Prep work
## 1.1 Prepare environment
```{r setup, include=FALSE, echo=FALSE}
library(mgcv)
library(mgcViz)
library(patchwork)
library(tidymv)
library(vegan)
library(ggvegan)
library(ggrepel)
library(ggforce)
library(tidyverse)
library(lubridate)
```

@@

weighted_fl_abund <- fl_abund_sp %>%
  group_by(site, date, plant.genus) %>%
  summarize(flower.cover = sum(fl.abund)) %>%
  left_join(weights) %>%
  mutate(weighted.flower.cover = flower.cover*visited) %>%
  group_by(site, date, bb.sp) %>%
  summarize(total.weighted.flower.cover = sum(weighted.flower.cover)) %>%
  left_join(site_data) %>%
  select(site, date, elev.mean, bb.sp, total.weighted.flower.cover) %>%
  mutate(yday = yday(date),
         bb.sp = factor(bb.sp),
         site = factor(site),
         year = factor(year(date)))
```

@@

### 2.2.1 Calculate max and mean abundances within site and year
```{r}
# Get mean total weighted floral abundance by BB sp
weighted_fl_abund_mean <- weighted_fl_abund %>%
  group_by(site, bb.sp, year) %>%
  summarize(mean.fl.abund = mean(total.weighted.flower.cover),
            max.fl.abund = max(total.weighted.flower.cover))

@@

### 2.2.2 Preliminary visualizations
```{r}
## mean floral abundance
ggplot(abund_sum, aes(elev.mean, mean.fl.abund, color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8), se = FALSE) +
  #facet_wrap(~bb.sp, scales = "free_y") +
  geom_vline(xintercept = 1500)

## log(mean floral abundance)
ggplot(abund_sum, aes(elev.mean, log(mean.fl.abund), color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8), se = FALSE) +
  facet_wrap(~bb.sp) +
  geom_vline(xintercept = 1500)

## Max floral abundance)
ggplot(abund_sum, aes(elev.mean, max.fl.abund, color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8), se = FALSE) +
  facet_wrap(~bb.sp) +
  geom_vline(xintercept = 1500)

## Max BB abundance
ggplot(abund_sum, aes(elev.mean, max.bb.abund, color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8),  se = FALSE) +
  facet_wrap(~bb.sp, scales = "free_y") +
  geom_vline(xintercept = 1500)

## Max  abundance ~ mean BB abundance, to see how interchangeable they are.
## They are almost perfectly correlated, and I think max abund is more informative
ggplot(abund_sum, aes(mean.bb.abund, max.bb.abund, color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8),  se = FALSE) +
  facet_wrap(~bb.sp, scales = "free_y")

## Are max FL and mean FL similarly interchangeable? Yes.
ggplot(abund_sum, aes(mean.fl.abund, max.fl.abund, color = year)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8),  se = FALSE) +
  facet_wrap(~bb.sp, scales = "free_y")

## Max BB abundance ~ mean FL abundance
ggplot(abund_sum, aes(log(mean.fl.abund), max.bb.abund)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "gam", formula = y ~ s(x, k = 8),  se = FALSE) +
  facet_wrap(~bb.sp, scales = "free_y")
```

### 2.2.3 Abundance models
I'm choosing to use site as a random intercept effect in these models because I think that is the correct way to specify a model involving repeated observations (in this case, bumble bee species) within site. The effect of this decision is that the elevation smoother is less influenced by individual sites and converges more strongly toward the mean. This creates less exciting curves, but I think it is a more principled way to model the data and less prone to overfitting. I can have my cake and eat it too, though, by plotting the actual data over the GAM fit so that the strong patterns of certain sites are visible. Then I can talk about the conditional elevation effect and the importance of certain peaks that are likely to be idiosyncrasies of my system (i.e. it would be overfitting to call them elevation effects).

#### 2.2.3.1 Bumble bees
```{r}
# Call model
gam_max_bb_abund_00 <- gam(max.bb.abund ~ 
                             s(bb.sp, bs = "re") +
                             s(bb.sp.year, bs = "re") + # fixed intercept for bb.sp:year
                             s(elev.mean, by = bb.sp, k = 8), # smoother for elev by BB sp.
                           method = "REML",
                           family = "tw", # Tweedie distribution performs well
                           select = TRUE,
                           data = abund_sum) %>% getViz()

@@

check.gamViz(gam_max_bb_abund_00) # evaluate model
gam_max_bb_abund_00_sum <- summary(gam_max_bb_abund_00) # summarize model
print(plot(gam_max_bb_abund_00, allTerms = TRUE), pages = 1) # preliminary visualization

# Plot smooths and year effect on the link scale
## Individual plots
gers_abund <- plot(sm(gam_max_bb_abund_00, 3)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "gers")
hort_abund <- plot(sm(gam_max_bb_abund_00, 4)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "hort")
jone_abund <- plot(sm(gam_max_bb_abund_00, 5)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "jone")
mend_abund <- plot(sm(gam_max_bb_abund_00, 6)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mend")
mont_abund <- plot(sm(gam_max_bb_abund_00, 7)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mont")
muci_abund <- plot(sm(gam_max_bb_abund_00, 8)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "muci")
pasc_abund <- plot(sm(gam_max_bb_abund_00, 9)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pasc")
prat_abund <- plot(sm(gam_max_bb_abund_00, 10)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "prat")
psit_abund <- plot(sm(gam_max_bb_abund_00, 11)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "psit")
pyre_abund <- plot(sm(gam_max_bb_abund_00, 12)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pyre")
soro_abund <- plot(sm(gam_max_bb_abund_00, 13)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "soro")
telu_abund <- plot(sm(gam_max_bb_abund_00, 14)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "telu")
wurf_abund <- plot(sm(gam_max_bb_abund_00, 15)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "wurf")

## Arrange significant smooths on grid plot
grid_bb_abund <- gridPrint(gers_abund, mend_abund, mont_abund, muci_abund, 
                           pasc_abund, prat_abund, psit_abund, pyre_abund, 
                           soro_abund, telu_abund, wurf_abund, ncol = 4)

pdf("../output/gam_max_bb_abund_grid.pdf", width = 8, height = 6)
plot(grid_bb_abund)
dev.off()

png("../output/gam_max_bb_abund_grid.png", res = 300, width = 8, height = 6, units = "in")
plot(grid_bb_abund)
dev.off()
```

#### 2.2.3.2 Floral resources
```{r}
gam_mean_fl_abund_00 <- gam(mean.fl.abund ~
                              s(elev.mean, bb.sp, bs = "fs", k = 8) +
                              s(elev.mean, year, bs = "fs", k = 8),
                           method = "REML",
                           family = "tw",
                           select = TRUE,
                           data = abund_sum) %>% getViz()

check.gamViz(gam_mean_fl_abund_00) # evaluate model
gam_mean_fl_abund_00_sum <- summary(gam_mean_fl_abund_00) # summarize model
print(plot(gam_mean_fl_abund_00, allTerms = TRUE), pages = 1) # preliminary visualization

# Plot on the link scale
elev_effect <- plot(sm(gam_mean_fl_abund_00, 2)) + 
  l_fitLine(alpha = 0.6, size = 1) + 
  labs(x = "Elevation (m.s.l.)", y = "Conditional effect (link scale)", color = "Year")

bb.sp_effect <- plot(sm(gam_mean_fl_abund_00, 1)) +
  l_fitLine(alpha = 0.6, size = 1) + 
  labs(x = "Elevation (m.s.l.)", y = NULL, color = "Species")

grid_fl_abund <- gridPrint(elev_effect, bb.sp_effect, ncol = 2)

pdf("../output/gam_mean_fl_abund_grid.pdf", width = 7, height = 3.5)
plot(grid_fl_abund)
dev.off()

png("../output/gam_mean_fl_abund_grid.png", res = 300, units = "in", width = 7, height = 3.5)
plot(grid_fl_abund)
dev.off()
```

>>>>

# 1. Prep work
## 1.1 Prepare environment
```{r setup, include=FALSE, echo=FALSE}
library(mgcv)
library(mgcViz)
library(gratia)
library(patchwork)
library(tidymv)
library(vegan)
library(ggvegan)
library(ggrepel)
library(ggforce)
library(tidyverse)
library(lubridate)
```

@@

weighted_flfl_abund <- fl_abund_sp %>%
  group_by(site, date, plant.genus) %>%
  summarize(flower.cover = sum(fl.abund)) %>%
  left_join(weights) %>%
  mutate(weighted.flower.cover = flower.cover*visited) %>%
  group_by(site, date, bb.sp) %>%
  summarize(total.weighted.flower.cover = sum(weighted.flower.cover)) %>%
  left_join(site_data) %>%
  select(site, date, elev.mean, bb.sp, total.weighted.flower.cover) %>%
  mutate(yday = yday(date),
         bb.sp = factor(bb.sp),
         site = factor(site),
         year = factor(year(date)))
```

@@

### 2.2.1 Calculate max and mean abundances within site and year
```{r}
# Get mean total weighted floral abundance by BB sp
weighted_fl_abund_mean <- weighted_flfl_abund %>%
  group_by(site, bb.sp, year) %>%
  summarize(mean.fl.abund = mean(total.weighted.flower.cover),
            max.fl.abund = max(total.weighted.flower.cover))

@@

### 2.2.2 Call models
```{r}
# BB
gam_max_bb_abund_00 <- gam(max.bb.abund ~ 
                             s(bb.sp, bs = "re") + # random intercept for bb.sp
                             s(bb.sp.year, bs = "re") + # random intercept for bb.sp:year
                             s(elev.mean, by = bb.sp, k = 8), # smoother for elev by bb.sp.
                           method = "REML",
                           family = "tw", # Tweedie distribution performs well
                           select = TRUE,
                           data = abund_sum) %>% getViz()

# FL
gam_mean_fl_abund_00 <- gam(mean.fl.abund ~
                              year +
                              s(bb.sp, bs = "re") +
                              s(elev.mean, by = bb.sp, k = 8),
                           method = "REML",
                           family = "tw",
                           select = TRUE,
                           data = abund_sum) %>% getViz()
```

### 2.2.3 Evaluate model
```{r}
# BB
check.gamViz(gam_max_bb_abund_00) # evaluate model
gam_max_bb_abund_00_sum <- summary(gam_max_bb_abund_00) # summarize model
print(plot(gam_max_bb_abund_00, allTerms = TRUE), pages = 1) # preliminary visualization

# FL
check.gamViz(gam_mean_fl_abund_00) # evaluate model
gam_mean_fl_abund_00_sum <- summary(gam_mean_fl_abund_00) # summarize model
print(plot(gam_mean_fl_abund_00, allTerms = TRUE), pages = 1) # preliminary visualization
```

### 2.2.4 Visualize

#### 2.2.4.1 BB
```{r}
# Plot smooths and year effect on the link scale

## Individual plots
gers_abund <- plot(sm(gam_max_bb_abund_00, 3)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "gers")
hort_abund <- plot(sm(gam_max_bb_abund_00, 4)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "hort")
jone_abund <- plot(sm(gam_max_bb_abund_00, 5)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "jone")
mend_abund <- plot(sm(gam_max_bb_abund_00, 6)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mend")
mont_abund <- plot(sm(gam_max_bb_abund_00, 7)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mont")
muci_abund <- plot(sm(gam_max_bb_abund_00, 8)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "muci")
pasc_abund <- plot(sm(gam_max_bb_abund_00, 9)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pasc")
prat_abund <- plot(sm(gam_max_bb_abund_00, 10)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "prat")
psit_abund <- plot(sm(gam_max_bb_abund_00, 11)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "psit")
pyre_abund <- plot(sm(gam_max_bb_abund_00, 12)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pyre")
soro_abund <- plot(sm(gam_max_bb_abund_00, 13)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "soro")
telu_abund <- plot(sm(gam_max_bb_abund_00, 14)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "telu")
wurf_abund <- plot(sm(gam_max_bb_abund_00, 15)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "wurf")

## Arrange smooths on grid plot

grid_bb_abund <- gridPrint(gers_abund, hort_abund, jone_abund, mend_abund, mont_abund, muci_abund, 
                           pasc_abund, prat_abund, psit_abund, pyre_abund, 
                           soro_abund, telu_abund, wurf_abund, ncol = 4)

pdf("../output/gam_bb_abund_grid.pdf", width = 8, height = 6)
plot(grid_bb_abund)
dev.off()

png("../output/gam_bb_abund_grid.png", res = 300, width = 8, height = 6, units = "in")
plot(grid_bb_abund)
dev.off()

# Extract predictions and plot on response scale

@@

gam_max_bb_abund_00_pred <- predict.gam(gam_max_bb_abund_00, # get model predictions
                                        type = "terms") %>% 
  as.data.frame() %>% # convert to data frame
  as_tibble() %>% # convert to tibble
  select(bb.sp = 1, bb.sp.year = 2, gers = 3, # rename
         hort = 4, jone = 5, mend = 6, mont = 7, 
         muci = 8, pasc = 9, prat = 10, psit = 11, 
         pyre = 12, soro = 13, telu = 14, wurf = 15) %>%
  mutate(fit = # get cumulative fit
           bb.sp + bb.sp.year + gers + 
           hort + jone + mend + mont + 
           muci + pasc + prat + psit + 
           pyre + soro + telu + wurf 
           ) %>%
  select(fit) %>%
  mutate(fit.trans = gam_max_bb_abund_00_ilink(fit)) %>% # back-transform to response scale
  bind_cols(abund_sum)

ggplot(gam_max_bb_abund_00_pred, aes(elev.mean, fit.trans, linetype = year)) + # map year to linetype
  geom_point(aes(elev.mean, max.bb.abund, color = tree_line), inherit.aes = FALSE, alpha = 0.5) + # add original data and map tree line to color
  geom_line(size = 0.5, alpha = 0.5) +
  facet_wrap(~bb.sp, scales = "free_y") +
  theme_light() +
  labs(x = "Elevation (m.s.l.)", 
       y = bquote("Peak abundance"),
       color = "Tree line",
       shape = "Year",
       linetype = "Year")

ggsave("../output/gam_bb_abund.pdf")
ggsave("../output/gam_bb_abund.png")
```

#### 2.2.4.2 FL
```{r}
# Plot on the link scale

## Individual plots

gers_fl_abund <- plot(sm(gam_mean_fl_abund_00, 2)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "gers")
hort_fl_abund <- plot(sm(gam_mean_fl_abund_00, 3)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "hort")
jone_fl_abund <- plot(sm(gam_mean_fl_abund_00, 4)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "jone")
mend_fl_abund <- plot(sm(gam_mean_fl_abund_00, 5)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mend")
mont_fl_abund <- plot(sm(gam_mean_fl_abund_00, 6)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "mont")
muci_fl_abund <- plot(sm(gam_mean_fl_abund_00, 7)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "muci")
pasc_fl_abund <- plot(sm(gam_mean_fl_abund_00, 8)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pasc")
prat_fl_abund <- plot(sm(gam_mean_fl_abund_00, 9)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "prat")
psit_fl_abund <- plot(sm(gam_mean_fl_abund_00, 10)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "psit")
pyre_fl_abund <- plot(sm(gam_mean_fl_abund_00, 11)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "pyre")
soro_fl_abund <- plot(sm(gam_mean_fl_abund_00, 12)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "soro")
telu_fl_abund <- plot(sm(gam_mean_fl_abund_00, 13)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "telu")
wurf_fl_abund <- plot(sm(gam_mean_fl_abund_00, 14)) + l_fitLine() + l_ciLine() + 
  labs(x = NULL, y = NULL, title = "wurf")
year_fl_effect <- plot(pterm(gam_mean_fl_abund_00, 1)) + l_fitPoints() + l_ciBar() + 
  labs(x = NULL, y = NULL, title = "Year")

## Arrange significant smooths on grid plot

grid_fl_abund <- gridPrint(gers_fl_abund, hort_fl_abund, jone_fl_abund, mend_fl_abund, mont_fl_abund, muci_fl_abund, 
                           pasc_fl_abund, prat_fl_abund, psit_fl_abund, pyre_fl_abund, 
                           soro_fl_abund, telu_fl_abund, wurf_fl_abund, year_fl_effect, ncol = 4)

pdf("../output/gam_fl_abund_grid.pdf", width = 8, height = 6)
plot(grid_fl_abund)
dev.off()

png("../output/gam_fl_abund_grid.png", res = 300, width = 8, height = 6, units = "in")
plot(grid_fl_abund)
dev.off()

# Extract predictions and plot on response scale

gam_mean_fl_abund_00_ilink <- family(gam_mean_fl_abund_00)$linkinv # get inverse link function to transform to the response scale

gam_mean_fl_abund_00_pred <- predict.gam(gam_mean_fl_abund_00, # get model predictions
                                        type = "terms") %>% 
  as.data.frame() %>% # convert to data frame
  as_tibble() %>% # convert to tibble
  select(year = 1, bb.sp = 2, gers = 3, # rename
         hort = 4, jone = 5, mend = 6, mont = 7, 
         muci = 8, pasc = 9, prat = 10, psit = 11, 
         pyre = 12, soro = 13, telu = 14, wurf = 15) %>%
  mutate(fit = # get cumulative fit
           year + bb.sp + gers + 
           hort + jone + mend + mont + 
           muci + pasc + prat + psit + 
           pyre + soro + telu + wurf 
           ) %>%
  select(fit) %>%
  mutate(fit.trans = gam_max_bb_abund_00_ilink(fit)) %>% # back-transform to response scale
  bind_cols(na.omit(abund_sum)) # the na.omit() drops the rows where floral surveying data were missing

ggplot(gam_mean_fl_abund_00_pred, aes(elev.mean, fit.trans, linetype = year)) + # map year to linetype
  geom_point(aes(elev.mean, mean.fl.abund, color = tree_line), inherit.aes = FALSE, alpha = 0.5) + # add original data and map tree line to color
  geom_line(size = 0.5, alpha = 0.5) +
  facet_wrap(~bb.sp, scales = "free_y") +
  theme_light() +
  labs(x = "Elevation (m.s.l.)", 
       y = bquote("Peak abundance"),
       color = "Tree line",
       shape = "Year",
       linetype = "Year")

ggsave("../output/gam_fl_abund.pdf")
ggsave("../output/gam_fl_abund.png")
```
